---
title: "FIrst Exploration"
author: "G.A. Pagani - M. Schipper"
date: "November 9, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
```

## Loading required libs.

```{r libs}
library(data.table)
library(ggplot2)
library(GGally)
library(reshape2)
library(chron)
Sys.setenv(TZ = "UTC")
```


Read data 
```{r readData}
data<-read.csv("data/dataframe2_10u55.csv")
```
Cleaning data and convert time and remove data where datetime is not available
Correct location of stations that have moved: 34 and 35 according to Bas excel sheet
```{r cleaning}
data$dateTime<-as.POSIXct(as.character(data$dateTime), format = "%Y-%m-%d %H:%M:%S")
data<-data.table(data)

data <- within(data, {
  dateTime[stn_ID==35 & dateTime>=as.POSIXct("2016-07-21 00:00:00", format = "%Y-%m-%d %H:%M:%S", tz="UTC") & 
             dateTime<= as.POSIXct("2016-08-04 00:00:00", format = "%Y-%m-%d %H:%M:%S", tz="UTC") ] <- NA
 
  stn_ID <- as.character(stn_ID)
  stn_ID[stn_ID=="34" &  dateTime<=as.POSIXct("2016-08-03 00:00:00", format = "%Y-%m-%d %H:%M:%S", tz="UTC")] <- "34a"
  stn_ID[stn_ID=="34"] <- "34b"
  stn_ID <- factor(stn_ID)
  NO2minusCAMS <- NO2-cams_NO2_3x3_min

})

 data<-data[!is.na(dateTime)]
```



```{r avgs}
#no effect this operation with the hourly aggregated data
avg.hr.no2<-data[,.(avg=mean(NO2)), by=.(YYYYMMDD,HH,stn_ID,dateTime)]
```



plot hourly time series

```{r plotTime, fig.width = 9, fig.height = 14}
ggplot(avg.hr.no2, aes(as.Date(dateTime),avg))+geom_line(aes(group = stn_ID))+ facet_wrap(~stn_ID, ncol = 3) +
  scale_x_date(date_labels = "%b") +
  labs(title = "Hourly average of NO2 concentration per location for 2016", y ="avg NO2", x = "calender time")
ggplot(avg.hr.no2, aes(as.Date(dateTime),avg))+geom_smooth(aes(group = stn_ID), method = "lm")+ facet_wrap(~stn_ID, ncol = 3) +
   scale_x_date(date_labels = "%b") +
  labs(title = "Linear trend of hourly average of NO2 concentration per location for 2016", y ="avg NO2", x = "calender time")
ggplot(avg.hr.no2, aes(as.Date(dateTime),avg))+geom_smooth(aes(group = stn_ID)) + facet_wrap(~stn_ID, ncol = 3) +
   scale_x_date(date_labels = "%b") +
    labs(title = "Smoothed trend of hourly average of NO2 concentration per location for 2016", y ="avg NO2", x = "calender time")


hour.24.cams<-data
ggplot(hour.24.cams, aes(as.Date(dateTime),cams_NO2_3x3_min))+geom_line(aes(group = stn_ID)) + facet_wrap(~stn_ID, ncol = 3) +
   scale_x_date(date_labels = "%b") 


```


```{r cleanStations}
avg.hr.no2<-subset(avg.hr.no2,!stn_ID %in% c(5,23))
data<-subset(data,!stn_ID %in% c(5,23))
```

```{r plotTime1}
ggplot(avg.hr.no2, aes(dateTime,avg))+geom_line(aes(group = stn_ID))+ facet_wrap(~stn_ID)
```

plot correlations

```{r correlationsHours}
test<-reshape(avg.hr.no2[,c(-1,-2)], idvar = "dateTime", timevar = "stn_ID", direction = "wide")
ggcorr(test[,-1],midpoint = .5, low=  "#EEEEEE")
```

```{r correlations with time}
test$dateTime <- as.numeric(test$dateTime)
ggcorr(test)
```


##24-Hourly
```{r average for each hour}
hour.24.avg<-data[,.(avg=mean(NO2)), by=.(HH,stn_ID)]
ggplot(hour.24.avg, aes(HH,avg))+geom_line(aes(group = stn_ID))+ facet_wrap(~stn_ID)
```

```{r correlations24Hours}
test2<-reshape(hour.24.avg, idvar = "HH", timevar = "stn_ID", direction = "wide")
ggcorr(test2[,-1],midpoint = .5, low=  "#EEEEEE")
```


### Weekdays vs. Weekend

```{r AddweekendIndicator}
data[,weekend:=is.weekend(dateTime)]
```

```{r weekendPlots , fig.width = 9, fig.height = 14}
hour.24.avg.weekend<-data[,.(avg=mean(NO2)), by=.(HH,stn_ID,weekend)]
ggplot(hour.24.avg.weekend, aes(HH,avg))+geom_line(aes(color=weekend))+ facet_wrap(~stn_ID) +
  scale_color_manual(name = "", values = 1:2, labels = c("mo-fr", "sa-su"))+
  labs(x = "hour of the day", y = "avg NO2", title = "weekend vs no weekend hourly average of NO2 concentration per location for 2016")
```
  
##adding CAMS
```{r CAMS}
avg.hr.cams<-subset(data,!stn_ID %in% c(5,23))
test.cams<-reshape(avg.hr.cams[,c("stn_ID","dateTime","cams_NO2_3x3_min")], idvar = "dateTime", timevar = "stn_ID", direction = "wide")
test.cams$dateTime <- as.numeric(test.cams$dateTime)
ggcorr(test.cams)
```

```{r CamsSame}
sum(test.cams$cams_NO2_3x3_min.12-test.cams$cams_NO2_3x3_min.35,na.rm = T)

```
correlations of the CAMS background
```{r camsCorr, fig.width = 9, fig.height = 9}
#avg.hr.cams[]
camsStation<-test.cams[,dateTime,cams_NO2_3x3_min.12]
camsStation[,stn_ID:="CAMS"]
camsStation[,dateTime:=as.POSIXct(camsStation$dateTime, origin="1970-01-01")]
setnames(camsStation,old = "cams_NO2_3x3_min.12","NO2")
#check time preservation?
test$cams<-camsStation$NO2
newnames <- gsub("avg." , "st_", names(test))
setnames(test, newnames)
ggcorr(test, midpoint = 0)+labs(title="Correlations between stations NO2 observations, CAMS, and time")
```

CAMS is added as an additional observation station
```{r cams as observation station}
camsObs<-melt.data.table(test,id.vars = "dateTime", variable.name = "stn_ID",value.name = "avg")
camsObs[,stn_ID:=gsub("avg.","",stn_ID)]
ggplot(camsObs, aes(dateTime,avg))+geom_line(aes(group = stn_ID))+ facet_wrap(~stn_ID)
ggplot(camsObs, aes(dateTime,avg))+geom_smooth(aes(group = stn_ID), method = "lm")+ facet_wrap(~stn_ID)
ggplot(camsObs, aes(dateTime,avg))+geom_smooth(aes(group = stn_ID)) + facet_wrap(~stn_ID)

```


visualization of data with CAMS contribution subtracted

```{r visualization cams subtracted}
ggplot(data, aes(dateTime,NO2minusCAMS))+geom_line(aes(group = stn_ID))+ facet_wrap(~stn_ID)
ggplot(data, aes(dateTime,NO2minusCAMS))+geom_smooth(aes(group = stn_ID), method = "lm")+ facet_wrap(~stn_ID)
ggplot(data, aes(dateTime,NO2minusCAMS))+geom_smooth(aes(group = stn_ID)) + facet_wrap(~stn_ID)
```



## Streets and Wind using NSL inner-product
# ```{r streets and wind}
# trafficAndRoads<-data[,dateTime,stn_ID,NO2,NO2minusCAMS,NSL_inproduct,NSL_INT_LV,NSL_INT_MV,NSL_INT_ZV,NSL_INT_BV,dist_SEGMENT_ID_nearest]
# 
# trafficAndRoads[,road_contr:=ifelse(NSL_inproduct>=0,yes = NSL_inproduct, no = 0)]
# ```





